---
title: "Data 1 Tutorial"
format: html
editor: visual
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
bp <- read_csv(here::here("data/bonita_peak.csv"))
```

# Overview

This is the quarto which runs the data 1 analysis on all the BP data

# Setup and Data

## Download and attach library

```{r}
#| eval: false
# libraries

# run this just once on each machine
# install.packages("tidyverse")

# load this EVERY time you start R
library(tidyverse)
```

## Data

-   Download data from D2L and put it in your `data/` folder.
-   Load the data into your RStudio session

```{r}
#| eval: false
# data

bp <- read_csv(here::here("data/bonita_peak.csv"))
```

### View data

```{r}
bp
```

### Filter data to two example sites

-   We will be just looking at two sites around Burrows Creek

::: callout-important
## Homework Note

-   You will be looking at two different sites in your Homework

-   Be sure to change the code in your homework script appropriately.
:::

```{r}
burrows <- bp |>
  filter(site_id =="BUaNFA" |
           site_id =="NFAaBU")
burrows
```

# Calculate Metrics

## Richness

```{r}
total_rich <- burrows |>
  distinct(stream, taxon) |>
  group_by(stream) |>
  summarise(total_rich = n())
total_rich
```

## E Richness

```{r}
E_rich <- burrows |>
  filter(order == "Ephemeroptera") |>
  distinct(stream, taxon) |>
  group_by(stream) |>
  summarise(EPT_rich = n())

E_rich
```

## EP Richness

```{r}
EP_rich <- burrows |>
  filter(order == "Ephemeroptera" |
           order == "Plecoptera") |>
  distinct(stream, taxon) |>
  group_by(stream) |>
  summarise(EPT_rich = n())

```

::: callout-important
## Homework Note

-   For your HW, you will calculate E, P, T richness separately, as well as the combined EPT richness
:::

## EPT Richness

```{r}
ept_rich <- burrows |>
  filter(order == "Ephemeroptera" |
           order == "Plecoptera" |
           order == "Trichoptera") |>
  distinct(stream, taxon) |>
  group_by(stream) |>
  summarise(EPT_rich = n())

```

# % EPT individuals

```{r}
ept_per_n <- burrows |>
  filter(order == "Ephemeroptera" |
           order == "Plecoptera" |
           order == "Trichoptera") |>
  group_by(stream, total_count) |>
  summarize(ept_n = sum(count)) |>
  mutate(ept_percent_ind = ept_n / total_count)
```

::: callout-important
## Homework Note

-   For your HW, you will also calculate the percent Chironomidae individuals.
:::

# % Chironomidae individuals

```{r}
chiro_per_n <- burrows |>
  filter(family == "Chironomidae") |>
  group_by(stream, total_count) |>
  summarize(chiro_n = sum(count)) |>
  mutate(chiro_percent_ind = chiro_n / total_count)
```

## Shannon

$$
H = - \Sigma \, p_i \, log \, p_i
$$

Where $p_i$ is the proportion of taxon $i$

```{r}
shannon <- burrows |> 
  select(stream, taxon, count, total_count) |>
  mutate(pi = count / total_count,
         ln_pi = log(pi),
         pi_ln_pi = pi*ln_pi) |>
  group_by(stream) |>
  summarize(H = -sum(pi_ln_pi))
```

# Join Results

```{r}
results <- full_join(total_rich,
                     ept_rich) |>
  full_join(ept_per_n) |>
  full_join(chiro_per_n) |>
  full_join(shannon)
```

# view results for 4 sites

```{r}
results 

```

# Plot Results

```{r}
results |>
  ggplot(aes(x = stream,
             y = total_rich)) +
  geom_col() +
  labs(x = "Site", 
       y = "Taxa Richness",
       title = "Taxa Richness")
```

```{r}
results |>
  ggplot(aes(x = stream,
             y = H)) +
  geom_col() +
  labs(x = "Site", 
       y = "H",
       title = "Shannon Diversity")
```
