---
title: "Data 1 Tutorial"
format: html
editor: visual
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
bp <- read_csv(here::here("data/bonita_peak.csv"))
```

# Overview

This is the tutorial for Data Module 1. This document will present the information necessary to complete the first data assignment in ENVS337.

# Project setup

-   Make an ENVS337 project

-   Set "save work space image" to never

-   Change pane layout to match in class example

-   Make a `data/` and `plots/` subfolder

# Setup and Data

## Download and attach library

```{r}
#| eval: false
# libraries

# run this just once on each machine
# install.packages("tidyverse")

# load this EVERY time you start R
library(tidyverse)
```

## Data

-   Download data from D2L and put it in your `data/` folder.
-   Load the data into your RStudio session

```{r}
#| eval: false
# data

bp <- read_csv("data/bonita_peak.csv")
```

### View data

```{r}
bp
```

## Data background

-   Mountain Studies Institute (MSI) in Durango/Silverton conducted this study

-   Sampled streams throughout the Bonita Peak Mining District located in San Juan County, CO

    -   [USEPA website](https://www.epa.gov/bpmd)

    -   [CDPHE website](https://cdphe.colorado.gov/hm/bonita-peak-mining-district)

-   Samples included benthic macroinvertebrates as well as water quality data

-   We will look at the macroinvertebrate data here

-   The water quality data will be incorporated in data module 2

## Data structure

-   Each row is the information for one taxon

-   Includes the `stream` name and specific `site` information as well as a unique `site_id`

-   There are columns for the complete taxonomic information for each taxon: `phylum`, `class`, `order`, `family`, and `genus`.

    -   When information is unknown it appears as an `NA` which means "Not Available"

-   The `taxon` column includes the lowest taxonomic information available for that group, typically a Genus name.

-   When the species information is known, the Genus and specific epithet are separated with an underscore

    -   i.e., *Drunella doddsii* appears as" `Drunella_doddssii`

-   There is also information on the `habit` 's and `FFG` 's for each taxon.

### Filter data to two example sites

-   We will be just looking at two sites around Burrows Creek

::: callout-tip
## Homework Note

-   You will be looking at two different sites in your Homework

-   Be sure to change the code in your homework script appropriately.
:::

```{r}
burrows <- bp |>
  filter(site_id =="BUaNFA" |
           site_id =="NFAaBU")
burrows
```

# Calculate Metrics

-   We will calculate a number of different metrics in class.

-   Some of these will be exactly the same on the homework assignment

-   Some of the homework problems will require you to modify the code to get the right answer.

## Total abundance

```{r}
burrows |>
  group_by(stream) |>
  distinct(total_count)
```

-   Note that this result appears in the console and then disappears

-   Below, we will save our results in a new data object for later use.

## Richness

```{r}
total_rich <- burrows |>
  distinct(stream, taxon) |>
  group_by(stream) |>
  summarise(total_rich = n())
total_rich
```

## EPT Richness

```{r}
ept_rich <- burrows |>
  filter(order == "Ephemeroptera" |
           order == "Plecoptera" |
           order == "Trichoptera") |>
  distinct(stream, taxon) |>
  group_by(stream) |>
  summarise(EPT_rich = n())
ept_rich
```

::: callout-tip
## Homework Note

-   For your HW, you will also calculate just the T richness

-   To do this, you will modify your `filter()` command to the following:

`filter(order == "Trichoptera") |>`
:::

## Heptageniidae Richness

-   Modify above code to filter based on the `family` column instead of the `order` column

```{r}
hept_rich <- burrows |>
  filter(family == "Heptageniidae") |>
  distinct(stream, taxon) |>
  group_by(stream) |>
  summarise(hept_rich = n())
hept_rich
```

-   Note that there are no Heptageniidae in Burrow creek so no line is returned.

# % EPT individuals

```{r}
ept_per_n <- burrows |>
  filter(order == "Ephemeroptera" |
           order == "Plecoptera" |
           order == "Trichoptera") |>
  group_by(stream, total_count) |>
  summarize(ept_n = sum(count)) |>
  mutate(ept_percent_ind = ept_n / total_count)
ept_per_n
```

::: callout-tip
## Homework Note

-   For your HW, you will also calculate the percent of individuals in the Chironomidae family
-   You will need to combine code from both the Heptageniidae and %EPT individuals examples above.
:::

## Scraper FFG

-   We will look at the scraper richness next.

-   But first, here is code to show you all the unique values in a column

```{r}
burrows |>
  distinct(FFG)
```

::: callout-tip
## Homework Note

-   Check the exact spelling of the FFG you want

-   The Homework will ask you to look at the richness of the predator FFG
:::

```{r}
scraper <- burrows |>
  filter(FFG == "scraper") |>
  distinct(stream, taxon) |>
  group_by(stream) |>
  summarise(hept_rich = n())
scraper
```

-   Note that once again there are no scrapers in Burrow creek, so there is no line in the result.

## Shannon's Diversity Index $H^\prime$

$$
H^\prime = - \Sigma \, p_i \, \ln(p_i)
$$

Where $p_i$ is the proportion of taxon $i$

```{r}
shannon <- burrows |> 
  select(stream, taxon, count, total_count) |>
  mutate(pi = count / total_count,
         ln_pi = log(pi),
         pi_ln_pi = pi*ln_pi) |>
  group_by(stream) |>
  summarize(H = -sum(pi_ln_pi))
```

# Plot Results

```{r}
 total_rich|>
  ggplot(aes(x = stream,
             y = total_rich)) +
  geom_col() +
  labs(x = "Site", 
       y = "Taxa Richness",
       title = "Taxa Richness")
```

-   You can also save this plot as a .png file in your R project

```{r}
#| eval: false
ggsave("plots/total_richness.png",
       units = "in",
       height = 4,
       width = 4)
```

-   Change the `height` and `width` columns as needed to get the right size

-   You will be pasting plots into your homework document.

-   Make sure they are the right size

-   Go back to your script and change the sizes if needed.

```{r}
shannon |>
  ggplot(aes(x = stream,
             y = H)) +
  geom_col() +
  labs(x = "Site", 
       y = "H",
       title = "Shannon Diversity")
```

::: callout-tip
## Homework Note

-   The homework will ask you to make 4 plots

-   Be sure to start with the object that has the result you want

    -   i.e.,`tot_rich`, `shannon`, etc.

-   In `aes()`, x will always be `x = stream`

-   You will need to change the y-axis to whatever you named the column in your object

    -   i.e., `y = total_rich`, `H`, etc.

-   Change the title and y label in the `labs()` argument to match your results.
:::

# Conclusion

-   This concludes the tutorial

-   You should now be able to calculate all the metrics in the first data assignment
