---
title: "MMI Assignment"
author: "Justin Pomeranz"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This homework assignment will be completed entirely in R.  

You will start a new R script which will have all of your code to answer the problems.  

If I ask for interpretation of your results, you can write out your answer directly in your script. Just be sure to start each line with a `#`. 

## Setup  

* Every programming script needs to be a stand-alone document  
* The first thing you should have is a title and author, and optionally a brief description of what the script does.  
* Next, You should load your libraries  
* Then load any data objects you will need  

* Your homework script should look something like this (be sure to change the relevant sections as needed )  

```{r, eval=FALSE}
# ENVS337 R assignment 2
# MMI
# YOUR NAME

# libraries
library(readxl)
library(tidyverse)

# load site data
site1 <- read_excel("data/MMI_HW_data.xlsx",
                    sheet = "A75CC")
site2 <- read_excel("data/MMI_HW_data.xlsx",
                    sheet = "A72")
other_sites <- read_excel("data/MMI_HW_data.xlsx",
                    sheet = "A72")

# Load taxa data
taxa_info <- read_excel("data/MMI_HW_data.xlsx",
                        sheet = "taxa_info")

# Load site data
site_info <- read_excel("data/MMI_HW_data.xlsx",
                        sheet = "site_info")

# combine site data
all_sites <- bind_rows(site1, site2, other_sites)

# add taxa info to site data
all_sites <- left_join(all_sites, taxa_info)

```

## Problem 1: pi_EPTnoB  

Calculate the score for the percent of individuals in EPT except in the Baetidae family.   

The following bullet points are an outline (in order) of the actions that need to occur to complete this calculation  

* First, group the data by site code by using `group_by(site_code)`  
* Use `mutate()` to calculate the total number of individuals in each site  
* Use `filter()` to **INCLUDE** EPT taxa  
* In the same `filter()` command, be sure to **EXCLUDE** the Baetidae family  
* Group your data by `site_code` and the total number of individuals per site (in the tutorial for "pi_SensitiveTax" this was named `tot_n`)  
* Use `summarize()` to sum the total number of individuals left in the data (i.e., `sum(Count))`  
* Use `mutate()` to calculate the metric (number of EPT_no_B / tot_n) and the score ($100 * metric / 79.3$)  
* use `select()` to only keep the `site_code` and the name of the column of the score you just created.  
* display your final results in the console  

## Problem 2: ClingerTax  

Calculate the score for the number of "Clinger" taxa.  

The following bullet points are an outline (in order) of the actions that need to occur to complete this calculation   

* `filter()` all the rows where `habit` is equal to "clinger"  
* Group your data by `site_code`  
* `summarize()` your data by counting the number of rows in each site  
  * i.e., use `n()` inside of summarize  
  
* use `mutate()` to calculate the score for this metric  
  * i.e. $100 * metric / 20$  
  
* use `select()` to only keep the `site_code` and the name of the column of the score you just created.  
* display your final results in the console  

## Problem 3: IntolerantTax  

Calculate the score for the number of "Intolerant" taxa.  

The following bullet points are an outline (in order) of the actions that need to occur to complete this calculation   

* `filter()` all the rows where `Bio2_sensitive` is equal to `TRUE`   
* Group your data by `site_code`  
* `summarize()` your data by counting the number of rows in each site  
  * i.e., use `n()` inside of summarize  
  
* use `mutate()` to calculate the score for this metric  
  * i.e. $100 * metric / 21$  
  
* use `select()` to only keep the `site_code` and the name of the column of the score you just created.  
* display your final results in the console  

## Problem 4: PredatorTax  

Calculate the score for the number of "predator" taxa.  

The following bullet points are an outline (in order) of the actions that need to occur to complete this calculation   

* `filter()` all the rows where `FFG` is equal to "predator"  
* Group your data by `site_code`  
* `summarize()` your data by counting the number of rows in each site  
  * i.e., use `n()` inside of summarize  
  
* use `mutate()` to calculate the score for this metric  
  * i.e. $100 * metric / 13$  
  
* use `select()` to only keep the `site_code` and the name of the column of the score you just created.  
* display your final results in the console  


## Problem 5: pi_ScraperTax  

Calculate the score for the percent of individuals that are in the "scraper" FFG.  

The following bullet points are an outline (in order) of the actions that need to occur to complete this calculation  

* First, group the data by site code by using `group_by(site_code)`  
* Use `mutate()` to calculate the total number of individuals in each site  
* Use `filter()` to include all the rows where `FFG` is equal to "scraper"    
* Group your data by `site_code` and the total number of individuals per site (in the tutorial for "pi_SensitiveTax" this was named `tot_n`)  
* Use `summarize()` to sum the total number of individuals left in the data (i.e., `sum(Count))`  
* Use `mutate()` to calculate the metric (number of scraper individuals / tot_n) and the score ($100 * metric / 44.2$)  
  
* use `select()` to only keep the `site_code` and the name of the column of the score you just created.  
* display your final results in the console  

## Problem 6: Metrics from Tutorial  

Copy the appropriate code from your notes/ the tutorial to calculate and display the scores for the following three metrics:  
1) TotalTax  
2) aEPTTax  
3) pi_SensitiveTax  

* Be sure to only include the `site_code` and whatever the column name that your score is stored in  

## Problem 7: Mean MMMI and site info  

Calculate the mean MMI score for each site.

* First, you will need to `left_join()` all of the results for each individual metric score  
* i.e., there should be 7 `left_join()` commands in sequence  
* once all your data is joined, calculate the mean MMI score  
* This was described in the "Averaging the MMI scores" section of the MMI Tutorial  

Hint:
* Use `rowwise()`  
* inside of `mutate()`, use the `c_across()` function with the names of your first and last column to be averaged separated with a `:`  
* `select()` the `site_code` and the name of the column which has your averages stored in it  

* Once your averages are calculated, use one more `left_join()` command to add the `site_info` to your results  
* Display your final object in the console here  

## Problem 8: Pairwise Plots  

* Produce a bar chart plot with all the appropriate pairwise combinations  
* i.e., you should be able to copy the code from the MMI tutorial  
* Just make sure to use the correct data object in the first line  

* Interpret your plot  
* In each pair, is the "reference" stream in attainment?  
* Is the "impacted" stream out of attainment?  
* Are any of the sites in the "Gray zone"?  
* based on these results, what would you say about the overall impact of the contamination spill in the Animas River?


## Problem 9: Longitudinal Plots   



### This concludes this R-assignment  


