---
title: "Single Metric Assignment"
author: "Justin Pomeranz"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This homework assignment will be completed entirely in R.  

You will start a new R script which will have all of your code to answer the problems.  

If I ask for interpretation of your results, you can write out your answer directly in your script. Just be sure to start each line with a `#`. 

## Setup  

* Every programming script needs to be a stand-alone document  
* The first thing you should have is a title and author, and optionally a brief description of what the script does.  
* Next, You should load your libraries  
* Then load any data objects you will need  

* Your homework script should look something like this (be sure to change the relevant sections as needed )  

```{r, eval=FALSE}
# ENVS337 R assignment 1
# Single metrics
# YOUR NAME

# libraries
library(readxl)
library(tidyverse)

# load data
site1 <- read_excel("data/MMI_HW_data.xlsx",
                    sheet = "A75CC")
site2 <- read_excel("data/MMI_HW_data.xlsx",
                    sheet = "A72")
taxa_info <- read_excel("data/MMI_HW_data.xlsx",
                        sheet = "taxa_info")
site_info <- read_excel("data/MMI_HW_data.xlsx",
                        sheet = "site_info")
```


## Problem 1: Single-site Richness Metrics  

For all sub-problems in problem 1, you will only be using the `site1` data object. 

#### Problem 1.1  
* Calculate the richness (number of taxa) which are in the Order Plecoptera  
* Save this result as a data object and print it to the console. 

#### Problem 1.2  
* Calculate the richness (number of taxa) which are in the Order Trichoptera  
* Save this result as a data object  
* join this result with the result calculated in 1.1 above and print it to the console. 

#### Problem 1.3  
* Calculate the richness (number of taxa) which are in the scraper FFG  
* Save this result as a data object  
* join this result with the results calculated above and print it to the console. 

## Problem 2: Single-site Metrics with individuals  

For all sub-problems in problem 2, you will only be using the `site1` data object. 

#### Problem 2.1  

* Calculate the percent of individuals which are in the Chironomidae Family  
* Save this result as a data object and print it to the console. 

#### Problem 2.2  

* Calculate the percent of individuals which are in the Scraper FFG  
* Save this result as a data object.  
* join this result with the result calculated in 2.1 above and print it to the console. 


## Problem 3: Multiple-site Metrics  

All sub-problems in problem 3 will work with the `site1` and `site2` data objects which have been combined using `bind_rows()`.  

```{r, eval=FALSE}
two_sites <- bind_rows(site1, site2)
```


#### Problem 3.1  

* Calculate the EPT richness for your two sites  
* Save this result as a new data object and print it to the screen  

#### Problem 3.2  

* Calculate the per cent of individuals which are in the shredder FFG. 
* Save this result as a new data object
* Join it to your result from 3.1 and print it to the screen  

#### Problem 3.3  

* Calculate the Shannon Diversity for these two sites.  
* Save this result as a new data object
* Join it to your previous results from problem 3 and print it to the screen  

#### Problem 3.4  

* Calculate the percent of individuals which are in the Chironomidae Family  
* Save this result as a new data object
* Join it to your previous results from problem 3 and print it to the screen  

#### Problem 3.5 Challenge (optional)   

This sub-problem is optional. Problem number 4 is not optional 

* Calculate the percent of individuals which are in the EPT orders but remove all individuals in the Baetidae Family    
* Save this result as a new data object
* Join it to your previous results from problem 3 and print it to the screen  


## Problem 4: Plots    
* All sub-problems in problem 4 will be working with the combined results from problem 3. i.e., the results from the two sites.   
* All plots made in part 4 need to have the following:  
  * A short, informative title.  
  * x- and y-axis labels  
  * Fill colors need to be based on the impact category variable.  
  * The color palettes that you use are up to you  

* Before you can plot, you will need to join the `site_info` to your results using the `left_join()` function. 
  
#### Problem 4.1  

* Make a bar plot of the EPT richness results  

#### Problem 4.2  

* Make a bar plot of the percent of individuals in the shredder FFG 

#### Problem 4.3  

* Make a bar plot of the Shannon Diversity results  

#### Problem 4.4  

* Make a bar plot of the percent of individuals in the Chironomidae Family results  

#### Problem 4.5 (Optional, if you did problem 3.5)  

* Make a bar plot of the percent of individuals in EPT orders without individuals which are in Baetidae results  

### This concludes this R-assignment  


