---
title: "Key Data 1"
subtitle: "Fall 2026"
execute: 
  echo: true
  eval: true
  message: false
  warning: false
format: html
editor: visual
toc: true
toc-depth: 2
embed-resources: true
theme:
  light: flatly
  dark: darkly
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
bp <- read_csv(here::here("data/bonita_peak.csv"))
```

## Overview

You will calculate some simple metrics for two sites from the Bonita Peak data set. The calculations will largely follow the tutorial from class. In addition to your calculations, you will need to interpret the results and answer some questions.

## Set up

-   Load the `tidyverse` library

-   Read in the data

```{r}
#| eval: false
# library
library(tidyverse)

# data
bp <- read_csv(here::here("data/bonita_peak.csv"))
```

## Modify data to keep two different sites

-   make an object called `data` which only has data from the following two `site_id` values

    -   `"MFMaM"`

    -   `"MAGaA"`

# Calculations

## Richness

1.  Calculate the total richness in your two sites. Make sure to save it as a data object.
