---
title: "Single Metric Assignment"
author: "Justin Pomeranz"
subtitle: "SOLUTION KEY"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = knitr::is_html_output())

library(readxl)
library(tidyverse)

site1 <- read_excel(here::here("data/MMI_HW_data.xlsx"),
                    sheet = "A75CC")
site2 <- read_excel(here::here("data/MMI_HW_data.xlsx"),
                    sheet = "A72")
taxa_info <- read_excel(here::here("data/MMI_HW_data.xlsx"),
                        sheet = "taxa_info")

site_info <- read_excel(here::here("data/MMI_HW_data.xlsx"),
                        sheet = "site_info")
site1 <- left_join(site1, taxa_info)
site2 <- left_join(site2, taxa_info)
```

## This is the KEY  


## Problem 1  
### Single-site Richness Metrics  

For all sub-problems in problem 1, you will only be using the `site1` data object. 

#### Problem 1.1  
* Calculate the richness (number of taxa) which are in the Order Plecoptera  
* Save this result as a data object and print it to the console. 

```{r}
p_rich <- site1 |>
  filter(Order == "Plecoptera") |>
  distinct(site_code, Taxon) |>
  group_by(site_code) |>
  summarize(p_rich = n())
p_rich
```


#### Problem 1.2  
* Calculate the richness (number of taxa) which are in the Order Trichoptera  
* Save this result as a data object  
* join this result with the result calculated in 1.1 above and print it to the console.

```{r}
t_rich <- site1 |>
  filter(Order == "Trichoptera") |>
  distinct(site_code, Taxon) |>
  group_by(site_code) |>
  summarize(t_rich = n())
result1 <- left_join(p_rich, t_rich)
result1
```

#### Problem 1.3  
* Calculate the richness (number of taxa) which are in the scraper FFG  
* Save this result as a data object  
* join this result with the results calculated above and print it to the console. 

```{r}
scrap_rich <- site1 |>
  filter(FFG == "scraper") |>
  distinct(site_code, Taxon) |>
  group_by(site_code) |>
  summarize(scrap_rich = n())
result1 <- left_join(result1, scrap_rich)
result1
```

## Problem 2  
### Single-site Metrics with individuals  

For all sub-problems in problem 2, you will only be using the `site1` data object. 

#### Problem 2.1  

* Calculate the percent of individuals which are in the Chironomidae Family  
* Save this result as a data object and print it to the console. 

```{r}
c_prop <- site1 |>
  select(site_code, Family, Count) |>
  mutate(tot_n = sum(Count)) |>
  filter(Family == "Chironomidae") |>
  group_by(site_code, tot_n) |>
  summarize(c_n = sum(Count)) |>
  mutate(c_prop = c_n / tot_n) |>
  select(site_code, c_prop)
c_prop
```

#### Problem 2.2  

* Calculate the percent of individuals which are in the Scraper FFG  
* Save this result as a data object.  
* join this result with the result calculated in 2.1 above and print it to the console. 

```{r}
scrap_prop <- site1 |>
  select(site_code, FFG, Count) |>
  mutate(tot_n = sum(Count)) |>
  filter(FFG == "scraper") |>
  group_by(site_code, tot_n) |>
  summarize(sc_n = sum(Count)) |>
  mutate(sc_prop = sc_n / tot_n) |>
  select(site_code, sc_prop)
result1.2 <- left_join(c_prop, scrap_prop)
result1.2
```

## Problem 3  
### Multiple-site Metrics  

All sub-problems in problem 3 will work with the `site1` and `site2` data objects which have been combined using `bind_rows()`.  

```{r, eval=TRUE}
two_sites <- bind_rows(site1, site2)
```


#### Problem 3.1  

* Calculate the EPT richness for your two sites  
* Save this result as a new data object and print it to the screen  

```{r}
ept_rich <- two_sites %>%
  filter(Order == "Ephemeroptera" |
           Order == "Plecoptera" |
           Order == "Trichoptera") |>
  distinct(site_code, Taxon) |>
  group_by(site_code) |>
  summarize(EPT_rich = n())
ept_rich
```


#### Problem 3.2  

* Calculate the per cent of individuals which are in the shredder FFG. 
* Save this result as a new data object
* Join it to your result from 3.1 and print it to the screen  

```{r}
shred_i <- two_sites %>%
  group_by(site_code) |>
  select(site_code, FFG, Count) |>
  mutate(tot_n = sum(Count)) |>
  filter(FFG == "shredder") |>
  group_by(site_code, tot_n) |>
  summarize(shred_n = sum(Count)) |>
  mutate(shred_percent_ind = shred_n / tot_n) |>
  select(site_code, shred_percent_ind)
a3.2 <- left_join(ept_rich, shred_i)
a3.2
```


#### Problem 3.3  

* Calculate the Shannon Diversity for these two sites.  
* Save this result as a new data object
* Join it to your previous results from problem 3 and print it to the screen 

```{r}
shannon <- two_sites |> 
  select(site_code, Taxon, Count) |>
  group_by(site_code) |>
  mutate(tot_n = sum(Count),
         pi = Count / tot_n,
         ln_pi = log(pi),
         pi_ln_pi = pi*ln_pi) |>
  group_by(site_code) |>
  summarize(H = -sum(pi_ln_pi))
a3.3 <- left_join(a3.2, shannon)
a3.3
```


#### Problem 3.4  

* Calculate the percent of individuals which are in the Chironomidae Family  
* Save this result as a new data object
* Join it to your previous results from problem 3 and print it to the screen  

```{r}
c_ind <- two_sites|>
  select(site_code, Family, Count) |>
  group_by(site_code) |>
  mutate(tot_n = sum(Count)) |>
  filter(Family == "Chironomidae") |>
  group_by(site_code, tot_n) |>
  summarize(c_n = sum(Count)) |>
  mutate(c_percent_ind = c_n / tot_n) |>
  select(site_code, c_percent_ind)
a3.4 <- left_join(a3.3, c_ind)
a3.4
```


#### Problem 3.5 Challenge (optional)   

This sub-problem is optional. Problem number 4 is not optional 

* Calculate the percent of individuals which are in the EPT orders but remove all individuals in the Baetidae Family    
* Save this result as a new data object
* Join it to your previous results from problem 3 and print it to the screen  

```{r}
ept_no_b <- two_sites |>
  select(site_code, Order, Family, Count) |>
  group_by(site_code) |>
  mutate(tot_n = sum(Count)) |>
  filter(Order == "Ephemeroptera" |
           Order == "Plecoptera" |
           Order == "Trichoptera",
         Family != "Baetidae") |>
  group_by(site_code, tot_n) |>
  summarize(ept_n = sum(Count)) |>
  mutate(ept_no_b_percent_ind = ept_n / tot_n) |>
  select(site_code, ept_no_b_percent_ind)
a3.5 <- left_join(a3.4, ept_no_b)
a3.5
```


## Problem 4   

### Plots    

* All sub-problems in problem 4 will be working with the combined results from problem 3. i.e., the results from the two sites.   
* All plots made in part 4 need to have the following:  
  * A short, informative title.  
  * x- and y-axis labels  
  * Fill colors need to be based on the impact category variable.  
  * The color palettes that you use are up to you  

* Before you can plot, you will need to join the `site_info` to your results using the `left_join()` function. 

```{r}
results <- left_join(a3.5, site_info)
```

  
#### Problem 4.1  

* Make a bar plot of the EPT richness results  

```{r}
ggplot(results,
       aes(x = site_code,
           y = EPT_rich,
           fill = impact_category)) +
  geom_bar(stat = "identity") +
  labs(title = "EPT Taxa Richness", 
       x = "Site code",
       y = "Number of taxa") +
  scale_fill_viridis_d() +
  theme_bw()
```


#### Problem 4.2  

* Make a bar plot of the percent of individuals in the shredder FFG 

```{r}
ggplot(results,
       aes(x = site_code,
           y = shred_percent_ind,
           fill = impact_category)) +
  geom_bar(stat = "identity") +
  labs(title = "Shredder percent of individuals", 
       x = "Site code",
       y = "Percent") +
  scale_fill_viridis_d() +
  theme_bw()
```

#### Problem 4.3  

* Make a bar plot of the Shannon Diversity results  

```{r}
ggplot(results,
       aes(x = site_code,
           y = H,
           fill = impact_category)) +
  geom_bar(stat = "identity") +
  labs(title = "Shannon Diversity", 
       x = "Site code",
       y = "H") +
  scale_fill_viridis_d() +
  theme_bw()
```

#### Problem 4.4  

* Make a bar plot of the percent of individuals in the Chironomidae Family results  

```{r}
ggplot(results,
       aes(x = site_code,
           y = c_percent_ind,
           fill = impact_category)) +
  geom_bar(stat = "identity") +
  labs(title = "Chironomidae percent individuals", 
       x = "Site code",
       y = "Percent") +
  scale_fill_viridis_d() +
  theme_bw()
```

#### Problem 4.5 (Optional, if you did problem 3.5)  

* Make a bar plot of the percent of individuals in EPT orders without individuals which are in Baetidae results  

```{r}
ggplot(results,
       aes(x = site_code,
           y = ept_no_b_percent_ind,
           fill = impact_category)) +
  geom_bar(stat = "identity") +
  labs(title = "EPT percent individuals with no Baetidae", 
       x = "Site code",
       y = "Percent") +
  scale_fill_viridis_d() +
  theme_bw()
```

## This concludes this R-assignment  


