---
title: "Single Metric Assignment"
author: "Justin Pomeranz"
subtitle: "SOLUTION KEY"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = knitr::is_html_output())

library(readxl)
library(tidyverse)

site1 <- read_excel(here::here("data/MMI_HW_data.xlsx"),
                    sheet = "A75CC")
site2 <- read_excel(here::here("data/MMI_HW_data.xlsx"),
                    sheet = "A72")
taxa_info <- read_excel(here::here("data/MMI_HW_data.xlsx"),
                        sheet = "taxa_info")

site_info <- read_excel(here::here("data/MMI_HW_data.xlsx"),
                        sheet = "site_info")
site1 <- left_join(site1, taxa_info)
site2 <- left_join(site2, taxa_info)
```

## This is the KEY  


## Problem 1  
### Single-site Richness Metrics  

For all sub-problems in problem 1, you will only be using the `site1` data object. 

#### Problem 1.1  
* Calculate the richness (number of taxa) which are in the Order Plecoptera  
* Save this result as a data object and print it to the console. 

```{r}
p_rich <- site1 |>
  filter(Order == "Plecoptera") |>
  distinct(site_code, Taxon) |>
  group_by(site_code) |>
  summarize(p_rich = n())
p_rich
```


#### Problem 1.2  
* Calculate the richness (number of taxa) which are in the Order Trichoptera  
* Save this result as a data object  
* join this result with the result calculated in 1.1 above and print it to the console.

```{r}
t_rich <- site1 |>
  filter(Order == "Trichoptera") |>
  distinct(site_code, Taxon) |>
  group_by(site_code) |>
  summarize(t_rich = n())
result1 <- left_join(p_rich, t_rich)
result1
```

#### Problem 1.3  
* Calculate the richness (number of taxa) which are in the scraper FFG  
* Save this result as a data object  
* join this result with the results calculated above and print it to the console. 

```{r}
scrap_rich <- site1 |>
  filter(FFG == "scraper") |>
  distinct(site_code, Taxon) |>
  group_by(site_code) |>
  summarize(scrap_rich = n())
result1 <- left_join(result1, scrap_rich)
result1
```

## Problem 2  
### Single-site Metrics with individuals  

For all sub-problems in problem 2, you will only be using the `site1` data object. 

#### Problem 2.1  

* Calculate the percent of individuals which are in the Chironomidae Family  
* Save this result as a data object and print it to the console. 

```{r}
c_prop <- site1 |>
  select(site_code, Family, Count) |>
  mutate(tot_n = sum(Count)) |>
  filter(Family == "Chironomidae") |>
  group_by(site_code, tot_n) |>
  summarize(c_n = sum(Count)) |>
  mutate(c_prop = c_n / tot_n) |>
  select(site_code, c_prop)
c_prop
```

#### Problem 2.2  

* Calculate the percent of individuals which are in the Scraper FFG  
* Save this result as a data object.  
* join this result with the result calculated in 2.1 above and print it to the console. 

```{r}
scrap_prop <- site1 |>
  select(site_code, FFG, Count) |>
  mutate(tot_n = sum(Count)) |>
  filter(FFG == "scraper") |>
  group_by(site_code, tot_n) |>
  summarize(sc_n = sum(Count)) |>
  mutate(sc_prop = sc_n / tot_n) |>
  select(site_code, sc_prop)
result1.2 <- left_join(c_prop, scrap_prop)
result1.2
```

## Problem 3  
### Multiple-site Metrics  

All sub-problems in problem 3 will work with the `site1` and `site2` data objects which have been combined using `bind_rows()`.  

```{r, eval=TRUE}
two_sites <- bind_rows(site1, site2)
```


#### Problem 3.1  

* Calculate the EPT richness for your two sites  
* Save this result as a new data object and print it to the screen  

#### Problem 3.2  

* Calculate the per cent of individuals which are in the shredder FFG. 
* Save this result as a new data object
* Join it to your result from 3.1 and print it to the screen  

#### Problem 3.3  

* Calculate the Shannon Diversity for these two sites.  
* Save this result as a new data object
* Join it to your previous results from problem 3 and print it to the screen  

#### Problem 3.4  

* Calculate the percent of individuals which are in the Chironomidae Family  
* Save this result as a new data object
* Join it to your previous results from problem 3 and print it to the screen  

#### Problem 3.5 Challenge (optional)   

This sub-problem is optional. Problem number 4 is not optional 

* Calculate the percent of individuals which are in the EPT orders but remove all individuals in the Baetidae Family    
* Save this result as a new data object
* Join it to your previous results from problem 3 and print it to the screen  


## Problem 4  
### Plots    
* All sub-problems in problem 4 will be working with the combined results from problem 3. i.e., the results from the two sites.   
* All plots made in part 4 need to have the following:  
  * A short, informative title.  
  * x- and y-axis labels  
  * Fill colors need to be based on the impact category variable.  
  * The color palettes that you use are up to you  

* Before you can plot, you will need to join the `site_info` to your results using the `left_join()` function. 
  
#### Problem 4.1  

* Make a bar plot of the EPT richness results  

#### Problem 4.2  

* Make a bar plot of the percent of individuals in the shredder FFG 

#### Problem 4.3  

* Make a bar plot of the Shannon Diversity results  

#### Problem 4.4  

* Make a bar plot of the percent of individuals in the Chironomidae Family results  

#### Problem 4.5 (Optional, if you did problem 3.5)  

* Make a bar plot of the percent of individuals in EPT orders without individuals which are in Baetidae results  

## This concludes this R-assignment  


